### 2025-05-05 Chapters/chapter4.tex
changes

\textbf{Why Ensure Balanced Testing Data?}
\begin{itemize}
    \item Prevents bias toward majority classes.
    \item Ensures the model’s performance is fairly evaluated.
    \item Helps achieve reliable generalization across all sleep stages.
\end{itemize}

The figure below shows the normalized class distribution during testing. Each class maintains an equalized density, avoiding class imbalance. This confirms that the model’s evaluation is not biased toward any specific sleep stage.

\textbf{Sampling Density Plot Showing Balanced Class Distribution:}
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/sample distribution plot pdf.png}
    \caption{Balanced class distribution across the test set.}
\end{figure}

\subsection{Model Performance: Training vs Testing}

Figures below demonstrate the accuracy and loss curves during training and testing, providing a clear visual comparison of model performance:

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/accuracy plot.png}
    \caption{Accuracy Curve: Training vs Testing.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/loss plot.png}
    \caption{Loss Curve: Training vs Testing.}
\end{figure}

\subsection{Model Evaluation: Confusion Matrix}

Performance metrics are calculated to ensure a comprehensive evaluation of the model's performance across all classes. We use precision, recall, and F1-score to assess the model:

\[
\text{Precision} = \frac{TP}{TP + FP}
\]
\[
\text{Recall} = \frac{TP}{TP + FN}
\]
\[
\text{F1-Score} = \frac{2 \times \text{Precision} \times \text{Recall}}{\text{Precision} + \text{Recall}}
\]

These metrics ensure a balanced evaluation of model performance across all sleep stages.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/confusion matrix samples.png}
    \caption{Confusion Matrix showing model performance across sleep stages.}
\end{figure}

\subsection{Gradient Analysis: Training Progression}

Understanding model training dynamics:

\textbf{Early Training (Epochs 0-5):} High loss with accuracy starting to improve.  
\textbf{Mid Training (Epochs 5-15):} Loss steadily decreases with stable gradient flow.  
\textbf{Late Training (Epochs 15-20):} Accuracy plateaus with no severe overfitting.

\textbf{Conclusion:} The training process remains stable, with no vanishing or exploding gradients.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/3d gradient.png}
    \caption{Gradient 3D Surface: Training vs Validation Metrics.}
\end{figure}

\subsection{Performance Metrics: Precision, Recall, F1-Score}

Evaluating model performance across all classes using key metrics:

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/precision_plot.png}
    \caption{Precision Scores per Class.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/recall_plot.png}
    \caption{Recall Scores per Class.}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/f1_score_plot.png}
    \caption{F1 Scores per Class.}
\end{figure}

\subsection{Feature Importance Analysis with LIME}

Understanding the contribution of different channels to model predictions, we used LIME (Local Interpretable Model-agnostic Explanations) to analyze feature importance. The analysis showed that the \textbf{EMG submental} and \textbf{EEG Pz-Oz} channels contribute the most to predictions, while the \textbf{EOG horizontal} channel has minimal importance, indicating lower relevance for classification.

This insight helps optimize feature selection and improve model efficiency.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{img/feature importance chanels analysis.png}
    \caption{Feature Importance Analysis for 4 Channels.}
\end{figure}

>>>>

\textbf{Why Ensure Balanced Testing Data?}
\begin{itemize}
	\item Prevents bias toward majority classes.
	\item Ensures fair evaluation across all sleep stages.
	\item Enhances generalization reliability.
\end{itemize}

The figure below shows the normalized class distribution during testing, confirming balance and the absence of class imbalance.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/sample_distribution_plot_pdf.png}
	\caption{Balanced class distribution across the test set.}
\end{figure}

\subsection{Model Performance: Accuracy and Loss Trends}

Figures below illustrate the model's training vs testing accuracy and loss progression over epochs.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/accuracy_plot.png}
	\caption{Accuracy Curve: Training vs Testing.}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/loss_plot.png}
	\caption{Loss Curve: Training vs Testing.}
\end{figure}

\subsection{Model Evaluation: Confusion Matrix and Evolution Metrics}

To evaluate classification performance, we compute the confusion matrix and derive the following metrics:

\[
\text{Precision} = \frac{TP}{TP + FP}, \quad
\text{Recall} = \frac{TP}{TP + FN}, \quad
\text{F1-Score} = \frac{2 \cdot \text{Precision} \cdot \text{Recall}}{\text{Precision} + \text{Recall}}
\]

Additionally, we compute the evolution matrix \( E \) that captures the transition probabilities between predicted and true classes:

\[
E(i, j) = \frac{C(i, j)}{\sum_{k} C(i, k)}
\]

Where \( C(i, j) \) is the confusion matrix count for true class \( i \) predicted as class \( j \).

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/confusion_matrix_samples.png}
	\caption{Confusion Matrix showing model performance across sleep stages.}
\end{figure}

\subsection{Gradient Analysis Across Training}

Model training stability was assessed via gradient trends:

\begin{itemize}
	\item \textbf{Epochs 0–5:} High loss, gradual accuracy increase.
	\item \textbf{Epochs 5–15:} Stable gradient flow and decreasing loss.
	\item \textbf{Epochs 15–20:} Accuracy plateaus, no overfitting.
\end{itemize}

\textbf{Conclusion:} Training remained stable with no vanishing/exploding gradients.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/3d_gradient.png}
	\caption{Gradient 3D Surface: Training vs Validation Metrics.}
\end{figure}

\subsection{Performance Metrics: Class-wise Scores}

Key metrics computed per class to understand classification effectiveness:

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/precision_plot.png}
	\caption{Precision Scores per Class.}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/recall_plot.png}
	\caption{Recall Scores per Class.}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/f1_score_plot.png}
	\caption{F1 Scores per Class.}
\end{figure}

\subsection{Feature Importance Analysis using LIME}

We used LIME to interpret feature contributions. The results highlight:

\begin{itemize}
	\item \textbf{Most influential:} EMG submental, EEG Pz-Oz
	\item \textbf{Least influential:} EOG horizontal
\end{itemize}

This helps refine feature selection and optimize performance.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{img/paper3/Analysis Plots/feature_importance_chanels_analysis.png}
	\caption{Feature Importance Analysis for 4 Channels using LIME.}
\end{figure}

### 2025-05-05 Chapters/introduction.tex
changes

Sleep is an essential component of human health, contributing significantly to cognitive function, memory consolidation, and emotional regulation. Adequate and quality sleep ensures optimal brain performance and supports overall physical and mental well-being. Disruptions in sleep patterns can lead to various health issues, including impaired concentration, mood disorders, and chronic conditions such as cardiovascular diseases.

The classification of sleep stages is fundamental in understanding sleep architecture and diagnosing sleep disorders. Experts categorize sleep into five primary stages—N1, N2, N3, REM, and Wake—based on distinct patterns observed in brain activity. Each sleep stage is characterized by specific brainwave frequencies, which can be detected using electroencephalography (EEG) signals. Accurate classification of these stages is crucial for analyzing sleep quality and identifying abnormalities such as insomnia, sleep apnea, and narcolepsy.

Table~\ref{tab:sleep_stages} provides an overview of the frequency ranges associated with each sleep stage.

\begin{table}[h]
\centering
\caption{Frequency ranges of different sleep stages.}
\begin{tabular}{ll}
\toprule
\textbf{Sleep Stage} & \textbf{Frequency Range (Hz)} \\
\midrule
Wake (Beta waves)       & 12--30 \\
N1 (Light Sleep)        & 4--8  \\
N2 (Moderate Sleep)     & 4--6  \\
N3 (Deep Sleep, Delta)  & 0.5--4  \\
REM (Theta waves)       & 4--6  \\
\bottomrule
\end{tabular}
\label{tab:sleep_stages}
\end{table}

To classify sleep stages effectively, bio-signals from multiple physiological channels are analyzed. The primary sources of information for sleep stage classification are EEG (electroencephalogram), EMG (electromyogram), and EOG (electrooculogram) signals. These channels capture brain activity, muscle movement, and eye movemets, respectively, offering a comprehensive view of sleep patterns.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\linewidth]{img/Sleep EDF Data Visulization.png}
    \caption{Sleep EDF Data Visulization}
    \label{fig:enter-label}
\end{figure}

Despite the availability of multiple channels in the SleepEDF dataset, not all contribute equally to classification performance. To reduce model complexity without compromising accuracy, we selectively utilize four channels from the nine standard channels provided in the dataset. The selected channels are:

\begin{itemize}
    \item \textbf{EEG Fpz-Cz}: Captures frontal-to-central brain activity.  
    \item \textbf{EEG Pz-Oz}: Measures parietal-to-occipital brain activity.  
    \item \textbf{EMG submental}: Records muscle tone from the chin area.  
    \item \textbf{EOG horizontal}: Tracks horizontal eye movements.  
\end{itemize}

In this study, we adopt a graph-based modeling approach to capture spatial relationships between the selected channels. Additionally, we employ a Transformer encoder to effectively model temporal dependencies across sleep cycles. To facilitate a better understanding of the input data, we also present visualizations of bio-signals from all channels.

This approach aims to leverage the strengths of both spatial and temporal modeling techniques, resulting in a robust and efficient sleep stage classification framework.

>>>>

Sleep is a fundamental aspect of human health that greatly contributes to cognitive function, memory consolidation, and emotional regulation. Proper and quality sleep guarantees proper brain function and overall physical and mental health. Sleep disturbances result in several diseases, such as impaired concentration, mood disorders, and chronic conditions like cardiovascular diseases.

Sleep plays a critical function in sustaining physical and mental well-being. It is not a passive resting state but a sophisticated biological process crucial for memory consolidation, emotional homeostasis, metabolic health, and immune system function. Alterations in sleep quality or sleep patterns can reflect or lead to many health problems, such as insomnia, depression, cardiovascular disease, and neurodegenerative disorders. It is essential for clinicians and scientists to precisely assess sleep stages for the diagnosis of such conditions and the prescription of proper treatment plans.

Staging of the sleep stages is the basis of understanding the architecture of sleep and the diagnosis of sleep disorders. Sleep is divided into five major stages—N1, N2, N3, REM, and Wake—based on unique patterns seen in brain function. Every stage of sleep is associated with particular frequencies of brainwaves, which can be picked up using the technique of electroencephalography (EEG) signals. The precise staging is important to study the quality of sleep and to detect abnormalities like insomnia, sleep apnea, and narcolepsy.

\subsection{The Role of EEG in Sleep Analysis}

Electroencephalography (EEG) is among the major methods of tracking brain activity while asleep. EEG captures electrical activity created by neurons as they fire within the brain, recorded with electrodes applied to the scalp. These electrical signals are critical in distinguishing various stages of sleep since each stage has different patterns of brain wave activity. Whereas other physiological signals are indirect, EEG is a direct view into the electrical activity of the brain and hence the basis for research and diagnosis into sleep.

\subsection{Understanding the SleepEDF Dataset}

The Sleep-EDF (Sleep European Data Format) corpus, popular among researchers, comprises polysomnography (PSG) recordings taken from healthy subjects and mildly disordered sleep patients. Recordings are taken non-invasively and usually overnight in a laboratory setting. In the course of a session, several sensors are applied to the subject's body to record a number of different physiological signals such as EEG (electroencephalogram), EOG (electrooculogram), and EMG (electromyogram).

Specifically, the database contains two EEG channels, namely Fpz-Cz and Pz-Oz, which record frontal and parietal brain activity. Additionally, it contains EOG signals to record eye movements and EMG signals to record muscle activity, particularly around the chin region. The recordings are manually annotated by experienced sleep technicians using visual patterns and set guidelines to mark sleep stages. The last annotation is retained in a hypnogram — a time series plot that shows the changes between various stages of sleep throughout the night.

\subsection{Sleep Stages and Signal Patterns}

Human sleep is commonly divided into five stages: Wake (W), Non-Rapid Eye Movement (NREM) stages N1, N2, and N3, and Rapid Eye Movement (REM). Each stage is characterized by specific frequency patterns in EEG signals:

\begin{table}[H]
	\centering
	\caption{Brainwave Types and Their Characteristics in Sleep Staging}
	\resizebox{\textwidth}{!}{%
		\begin{tabular}{@{}lll@{}}
			\toprule
			\textbf{Wave Type} & \textbf{Frequency Range} & \textbf{Associated Sleep Stage / Behavior} \\ \midrule
			Alpha waves        & 8–13 Hz                 & Relaxed wakefulness, especially with eyes closed \\
			Beta waves         & 13–30 Hz                & Alertness and active thinking; decrease during sleep \\
			Theta waves        & 4–8 Hz                  & Light sleep (Stages N1 and N2) \\
			Delta waves        & 0.5–4 Hz                & Deep sleep (Stage N3); synchronized neuronal firing \\
			Sleep spindles \& K-complexes & ~12–15 Hz (spindles)     & Characteristic of Stage N2; used for stage classification \\ 
			\bottomrule
		\end{tabular}%
	}
	\label{tab:brainwaves}
\end{table}

These patterns are extracted from raw EEG signals through filtering and segmentation. Experts use these patterns, along with eye movement and muscle tone data, to classify each 30-second segment (epoch) into one of the five stages.

\subsection{The Need for Automation}
Manual scoring of sleep stages, while precise, is time-consuming and subject to inter-rater reliability. It takes hours of professional scoring for one night's worth of recording. Automated sleep stage scoring provides a quicker, more scalable, and more possibly consistent solution. Using machine learning and deep learning models, particularly those that can address intricate spatial and temporal patterns in physiological data, we can develop systems that emulate expert opinion and aid in clinical decision-making.
This thesis suggests just such a system — \textit{SleepGCN-Transformer} — that unifies graph-based representation of EEG sensor relationships with temporal learning from transformers. Our strategy not only targets high classification performance but also contributes to the increasing body of research in interpretable AI in medicine, moving us closer to clinically integrated, explainable, and fully automated sleep diagnostics.

### 2025-05-05 Chapters/paper3.tex
changes

We begin by importing the Sleep-EDF data provided in EDF (European Data Format). The raw EEG recordings are loaded using MNE's \texttt{read\_raw\_edf} function, and corresponding annotations are aligned using \texttt{read\_annotations}. We optionally exclude all non-EEG channels unless explicitly required, focusing on EEG signals for our analysis. The dataset includes two EEG channels: \texttt{Fpz-Cz} and \texttt{Pz-Oz}, which are the primary input sources in our work.

During loading, we crop the signal to reduce unnecessary wake-time data, retaining 30 minutes before the first sleep stage and 30 minutes after the last. This ensures that our training data remains focused around relevant sleep activity. The channel names are cleaned by stripping prefixes for standardization.

After loading and cropping, we segment the data into fixed-length 30-second epochs using the known sampling frequency of the recordings. Only epochs labeled with valid sleep stages — specifically stages W, N1, N2, N3, and REM — are retained. For each epoch, the raw data is extracted and structured into a format suitable for training.

Once all epochs are collected, we flatten the 3D EEG data (epochs $\times$ channels $\times$ time points) into 2D arrays (epochs $\times$ features). This is necessary for traditional machine learning classifiers. To address class imbalance in sleep stages, we apply SMOTE (Synthetic Minority Oversampling Technique) to generate balanced training samples.

The balanced dataset is then split into training and testing sets in an 80-20 ratio. We train a  Machine Learning   on the training data. After training, predictions are made on the test set, and the model is evaluated using standard metrics including classification report and confusion matrix. The confusion matrix is visualized using a heatmap to give a clear view of the stage-wise performance of the classifier.

@@

We have used the Sleep-EDF Dataset available at \url{https://www.physionet.org/content/sleep-edfx/1.0.0/}. For our experiments, we selected recordings from approximately 30 patients, resulting in 60 files — 30 EDF (European Data Format) recordings and their corresponding hypnogram annotation files. The EDF files include various bio-signals captured through multiple channels. The hypnogram files contain sleep stage annotations, specifying the start time and type of each stage (Wake, N1, N2, N3, N4, REM) with timestamps in the HH:MM:SS format.

The EDF recordings consist of a range of physiological signals, including EEG (\texttt{Fpz-Cz} and \texttt{Pz-Oz}), EOG, EMG (submental), rectal temperature, and oro-nasal respiration. Each channel provides continuous signal recordings at a specific sampling frequency (Hz), which are used as input features for our downstream machine learning pipeline. From these signals, we derive two essential inputs: spatial features, used to construct graph-based representations, and temporal features, which capture the sequence and timing dynamics necessary for modeling the sleep process.

\begin{figure}
	\centering
	\includegraphics[width=0.4\linewidth]{"img/paper 1/EEG_Filter_chanel"}
	\caption{Filter EEG Chanels}
	\label{fig:eegfilterchanel}
\end{figure}

\section{Preprocessing}

This study applied machine learning techniques to classify sleep stages using EEG signals. The preprocessing pipeline included several crucial steps: channel mapping, data cropping, signal filtering, epoch creation, label mapping, and class balancing using SMOTE. These steps ensured clean, well-structured input for feature extraction and model training. The extracted features spanned time-domain, frequency-domain, and time-frequency representations. Multiple machine learning models—Gradient Boosting, Random Forest, Support Vector Machine (SVC), and Bagging—were trained and evaluated using metrics such as accuracy, precision, recall, and F1-score with cross-validation to assess generalization. The trained models were finally used to classify unseen EEG data into five sleep stages: Wake, N1, N2, N3, and REM.

\paragraph{A. Dataset Acquisition}

EEG signals were recorded using non-invasive electrodes placed on the scalp of human volunteers, capturing brain electrical activity during overnight sleep studies. Additional demographic data such as age, gender, medical history, and sleep habits were collected to provide contextual understanding. The data collection took place in controlled conditions to minimize noise and interference. Manual inspection of the raw EEG was performed to reject or correct artifacts where necessary, ensuring the quality of the input data. The final recordings were saved in standard EDF format for compatibility with downstream analysis. The dataset used in this study was sourced from publicly available Sleep-EDF recordings, which include EEG data and associated hypnogram annotations reflecting sleep stages.

\paragraph{B. Data Preprocessing}

Preprocessing was applied to prepare EEG data for modeling. Channel mapping was done to identify relevant EEG channels, particularly \texttt{Fpz-Cz} and \texttt{Pz-Oz}, and irrelevant or noisy channels were excluded. Data cropping was applied to isolate the sleep period from wake periods based on annotation markers. The signals were then filtered to remove noise and isolate specific frequency bands important for sleep analysis.

Epochs of 30-second duration were created from continuous EEG signals to structure the data for machine learning. Sleep stage labels corresponding to these epochs were derived from hypnogram annotations. Given the naturally imbalanced class distribution in sleep stage data (e.g., fewer REM and N1 stages), the SMOTE algorithm was employed to oversample the minority classes, thereby enhancing model robustness. After preprocessing, the dataset was split into training and testing subsets. This enabled reliable training of models and unbiased performance evaluation. The output from preprocessing was then used in the feature extraction and modeling stages.

@@

The core modeling pipeline begins with transforming the raw EEG data into structured epochs suitable for machine learning classification. Each EEG recording is segmented into 30-second epochs using the sampling frequency and annotated sleep stages. Only valid sleep stages (Wake, N1, N2, N3, and REM) are retained for downstream processing. After epoching, each segment is extracted using the \texttt{mne.Epochs} function, resulting in structured EEG data with consistent time windows. These segments are then reshaped into two-dimensional feature vectors where each row corresponds to an epoch and each column to a time-series sample across all EEG channels.

To address the class imbalance inherent in sleep stage data, the SMOTE (Synthetic Minority Over-sampling Technique) algorithm is applied. This balances the dataset by synthetically generating new examples in underrepresented classes. The balanced dataset is then split into training and testing sets in an 80:20 ratio.

\begin{figure}
	\centering
	\includegraphics[width=0.7\linewidth]{"img/paper 1/Architechture1"}
	\caption{Architecture of Machine Learning Classifiers for Sleep Stage
		Classificatio}
	\label{fig:architechture1}
\end{figure}

\subsection{Random Forest Classifier}

We implemented a Random Forest classifier to establish a strong baseline model for sleep stage classification. The raw EEG recordings were segmented into 30-second epochs, and features were extracted by flattening the multidimensional epoch data. To handle class imbalance, the SMOTE algorithm was applied prior to training. The model was initialized with \textbf{100 decision trees} (\texttt{n\_estimators=100}) and a fixed \textbf{random seed} (\texttt{random\_state=42}) to ensure reproducibility. This configuration allowed the model to generalize well while maintaining robust performance across varying sleep stages.

\subsection{Gradient Boosting Classifier with PCA}

To improve computational efficiency and potentially boost model accuracy, we introduced Principal Component Analysis (PCA), retaining \textbf{95\% of the explained variance} (\texttt{n\_components=0.95}). The reduced feature set was passed to a \textbf{Gradient Boosting classifier} configured with \textbf{30 estimators} (\texttt{n\_estimators=30}), a \textbf{maximum tree depth of 3} (\texttt{max\_depth=3}), and a \textbf{learning rate of 0.1}. Class balancing was again addressed with SMOTE. The same 80-20 train-test split was maintained with \texttt{random\_state=42}. This model leverages the sequential learning of weak classifiers to optimize classification performance over multiple iterations.

\subsection{Ensemble Learning (Voting Classifier)}

\begin{sloppypar}
	For further enhancement, we employed a soft voting ensemble that combines multiple classifiers, each with complementary strengths. The ensemble includes a \textbf{Gradient Boosting Classifier} (\texttt{n\_estimators=50}, \texttt{max\_depth=3}, \texttt{learning\_rate=0.1}), a \textbf{Random Forest Classifier} (\texttt{n\_estimators=50}, \texttt{max\_depth=3}), a \textbf{Support Vector Classifier (SVC)} with an RBF kernel and probability estimates enabled, and a \textbf{Bagging Classifier} (\texttt{n\_estimators=30}, \texttt{max\_samples=1.0}, \texttt{max\_features=1.0}). These classifiers were integrated using a \textbf{soft voting strategy} in the \texttt{VotingClassifier(voting='soft')} to average their probabilistic predictions.

	Feature reduction was again performed using PCA with 95\% variance retention. This ensemble approach leverages model diversity to achieve improved generalization and more stable predictions across sleep classes.
\end{sloppypar}

@@

This section presents the results and evaluations for the models trained, including Random Forest, Bagging, Ensemble Learning, and Gradient Boosting. The following figures summarize the performance metrics for each model.

\subsection{Random Forest}

For the Random Forest model, we present the following evaluation metrics:

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/random_forest_confusion_matrix.png}
	\caption{Random Forest Confusion Matrix}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/accuracy_curve.png}
	\caption{Random Forest Accuracy Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/loss_curve.png}
	\caption{Random Forest Loss Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Random_forest/precision_per_class.png}
	\caption{Random Forest Precision Per Class}
\end{figure}

\subsection{Bagging Classifier}

For the Bagging Classifier model, we present the following evaluation metrics:

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/BG_confusion_matrix.png}
	\caption{Bagging Classifier Confusion Matrix}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/accuracy_curve.png}
	\caption{Bagging Classifier Accuracy Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/loss_curve.png}
	\caption{Bagging Classifier Loss Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Bagging_Boosting/precision_per_class.png}
	\caption{Bagging Classifier Precision Per Class}
\end{figure}

\subsection{Ensemble Learning}

For the Ensemble Learning model, we present the following evaluation metrics:

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/Ensemble_confusion_matrix.png}
	\caption{Ensemble Learning Confusion Matrix}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/accuracy_curve.png}
	\caption{Ensemble Learning Accuracy Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/loss_curve.png}
	\caption{Ensemble Learning Loss Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/Ensemble_Learning/precision_per_class.png}
	\caption{Ensemble Learning Precision Per Class}
\end{figure}

\subsection{Gradient Boosting}

For the Gradient Boosting model, we present the following evaluation metrics:

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/confusion_matrix.png}
	\caption{Gradient Boosting Confusion Matrix}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/accuracy_curve.png}
	\caption{Gradient Boosting Accuracy Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/loss_curve.png}
	\caption{Gradient Boosting Loss Curve}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{../img/paper_1/gradient_boosting/precision_per_class.png}
	\caption{Gradient Boosting Precision Per Class}
\end{figure}

>>>>

@@

\section{Preprocessing Techniques}

@@

@@

### 2025-05-05 Title/abstract.tex
changes

\begin{center}
\section*{\textcolor{internationalkleinblue}{\textbf{{Abstract}}}}
\end{center} 
This project presents \textit{SleepGCN-Transformer}, a hybrid model combining Graph Convolutional Networks (GCN) and Transformer encoders for sleep stage classification. Using the SleepEDF dataset, it incorporates four physiological signals: EEG (Fpz-Cz, Pz-Oz), EMG (submental), and EOG (horizontal). Preprocessing includes 30-second epochs, band-pass filtering (0.3--30 Hz), and graph-based EEG channel representation.

The GCN module captures spatial relationships across EEG channels, while the Transformer encoder models temporal dependencies from graph-level embeddings. Focal Loss addresses class imbalances, and a CosineAnnealingLR scheduler optimizes learning rate decay. Training with the AdamW optimizer for 20 epochs achieves 93.12 \% training and 93.04\% validation accuracy.

Model performance, assessed using precision, recall, and F1-score, demonstrates high efficacy. LIME-based feature importance analysis highlights EMG and EEG Pz-Oz channels as key contributors. This hybrid model exhibits state-of-the-art performance, with future directions focused on enhancing explainability and clinical integration.

>>>>

\begin{center}
	\section*{\textcolor{internationalkleinblue}{\textbf{{Abstract}}}}
\end{center}

Correct and effective sleep stage classification is crucial in the diagnosis of sleep disorders, conventionally done through manual interpretation of polysomnography (PSG) signals. Automated sleep stage classification is investigated in this thesis by a series of increasingly sophisticated models, from traditional machine learning to sophisticated deep learning architectures. Initial trials on the SleepEDF dataset compared models like Random Forest, XGBoost, and ensemble models, which obtained maximum accuracy of 85.35\%, and set the foundation for automated systems to perform consumer-level sleep monitoring. Future work introduced deep learning models like BiLSTM and RNNs, and BiLSTM reached 81.13\% accuracy, which proved to have better ability to model temporal sequences.

Based on these building blocks, we introduce \textit{SleepGCN-Transformer}, a new hybrid model that combines Graph Convolutional Networks (GCNs) to learn spatial relations between multi-channel EEG, EMG, and EOG signals, and Transformer encoders to learn temporal patterns. The model applies Focal Loss to handle class imbalance and a CosineAnnealingLR scheduler for dynamic learning rate adaptation. Trained with the AdamW optimizer on preprocessed 30-second epochs, SleepGCN-Transformer is 93.12\% training and 93.04\% validation accurate. Feature attribution by LIME indicates EMG and EEG Pz-Oz to be the most significant channels, demonstrating the model's interpretability.

### 2025-05-05 Title/title.tex
changes

\begin{center}
    {{\Large \textbf{SleepGCN-Transformer: A Hybrid Graph Convolutional and
Transformer Network for Sleep Stage Classification
}}}\\
    \vspace{0.6cm}
    {\large A Master's Report}\\
    {\large Submitted in partial fulfillment of the requirement for the award of the degree of}\\
    \vspace{0.8cm}
    {\large Master of Technology in Artificial Intelligence}\\
    {\large by}\\
    \textbf{{\Large  Tanmay Rathod}}\\
    {\large 23MAI007}\\
    \vspace{0.7cm}
    {\large Under the guidance of}\\
    {\large  \textbf{Dr. Santosh Kumar Satapathy}  {\large }}\\
    {\large Department of Information and Communication Technology}\\
    \vspace{0.6cm}
        \vfill
    {\centering \includegraphics[width=0.21\textwidth]{img/pdpu_logo.png}}\\
    {\large School of Technology}\\
    {\large Pandit Deendayal Energy University}\\
    {\large Gandhinagar – 382426. Gujarat - India}\\
    {\large February, 2025}
\end{center}

>>>>

\begin{center}
	{{\Large \textbf{SleepGCN-Transformer: A Hybrid Graph Convolutional and\\
				Transformer Network for Sleep Stage Classification}}}\\
	\vspace{0.6cm}

	{\large A Master's Thesis}\\
	{\large Submitted in partial fulfillment of the requirement for the award of the degree of}\\
	\vspace{0.8cm}

	{\large Master of Technology in Artificial Intelligence}\\
	{\large by}\\

	\textbf{{\Large Tanmay Rathod}}\\
	{\large 23MAI007}\\
	\vspace{0.7cm}

	{\large Under the guidance of}\\
	{\large \textbf{Dr. Santosh Kumar Satapathy}}\\

	{\large Department of Information and Communication Technology}\\
	\vspace{0.6cm}

	\vfill
	{\centering \includegraphics[width=0.21\textwidth]{img/pdpu_logo.png}}\\

	{\large School of Technology}\\
	{\large Pandit Deendayal Energy University}\\
	{\large Gandhinagar – 382426. Gujarat - India}\\
	{\large May, 2025}
\end{center}
